Integrating Common Ground and Informativeness in Pragmatic Word Learning

### 2019-01-28 papers/cogsci2019/cogsci2019.tex
cogsci draft

\begin{abstract}
Pragmatic inferences are an integral part of language comprehension and
learning. To recover the intended meaning of an utterance, listeners
need to balance and integrate different sources in information. In a
series of experiments, we studied how listeners integrate general
expectations about speakers with expectations that are specific to the
interactional history with a particular speaker. We used a Bayesian
cognitive model to formalize the integration process. In Experiment 1
and 2 we replicated findings showing that listeners make inferences
based on general and speaker specific expectations. Using the empirical
measurements from these experiments, we generated model predictions
about how they should be integrated. We tested these predictions in
Experiment 3, which combined both expectations in a single design.
Experiment 4 replicated and extended Experiment 3 to a broader set of
conditions. Across both experiments, listeners based their inferences on
both types of expectations. Overall, model predictions and data were
highly correlated. Predictions from the model incorporating both types
of information described listeners' behavior better compared to
alternative models based on only one type of information. This research
shows that listeners flexibly integrate different forms of social
expectations across a range of contexts, a process which can be
described using Bayesian cognitive models. Our approach highlights how
experiments and computational models can be used in conjunction to
inform theories about language use and comprehension.

@@

One of the most astonishing features of natural language is that it
allows us to communicate precise meanings despite the fact that each
utterance is inherently ambiguous. The conventional couplings between
sounds (words) and objects constrain what a speaker may mean by an
utterance. However, the intended meaning of the utterance is not
reducible to the words that are contained in it. For example, the same
utterance (``Look, there it is!'') could be intended to refer to very
different things depending on whether someone is searching for their
wallet or looking for the bus stop. It takes additional pragmatic
inferences to recover the intended meaning (Levinson, 2000).

Pragmatic inferences rest on a set of expectations that interlocutors
bring to the table when entering a communicative interaction. On the one
hand, speakers and listeners have the general expectation that their
partner communicates in an informative and relevant way (Sperber \&
Wilson, 2001). Grice (1991) summarised this expectation via the
cooperative principle: ``Make your contribution such as is required, at
the stage at which it occurs, by the accepted purpose or direction of
the talk exchange in which you are engaged.'' Importantly, the second
half of the principle highlights a second type of expectation:
interlocutors expect each other to produce and interpret utterances in
light of the shared common ground between them (H. H. Clark, 1996).
Common ground refers to bits of information that are assumed to be
shared, either because they result from joint experience or were
otherwise grounded (Bohn \& Koymen, 2018). The nature of the
corresponding expectations vary with the individuals involved.

Both general and common ground expectations can support children's word
learning (E. V. Clark, 2009; Tomasello, 2009). On the one hand, children
have been found to learn novel words by assuming that the speaker is
informative (Frank \& Goodman, 2014). That is, in the absence of any
prior interaction with the speaker, children interpreted a novel word as
referring to the most informative referent. On the other hand, children
learn novel words by relying on common ground expectations (Akhtar,
Carpenter, \& Tomasello, 1996; Bohn \& Koymen, 2018). For example, when
a speaker expressed preference for a particular object, children expect
a novel word from the same speaker to refer to the previously preferred
object (Saylor, Sabbagh, Fortuna, \& Troseth, 2009).

But how are general and common ground expectations integrated with one
another? Are pragmatic inferences strengthened additively when both
support a particular interpretation? How are they weighed when they are
in conflict? The rational speech act (RSA) framework (Frank \& Goodman,
2012) offers a formal framework for this information integration
problem. RSA models are characterized by their recursive structure in
which speakers and listeners reason about each other's interpretation of
utterances in context. RSA models have been found to make accurate
quantitative predictions about various forms of pragmatic language use
and word learning (Goodman \& Frank, 2016). However, a comprehensive
treatment of how general and common ground expectations are integrated,
is still missing.

Within RSA models, each agent in this recursion is modeled as a Bayesian
reasoner; thus, information integration is treated as a process of
probabilistic inference. A natural locus within these models for
information integration is the tradeoff between the prior probability of
a particular referent (the degree to which it is in common ground) and
the likelihood of that referent given the current utterance (general
pragmatic interpretation).

Here we investigate information integration in pragmatic reasoning by
experimentally manipulating speaker specific and common ground
information and testing how adult listeners integrate them in a word
learning scenario. In Experiment 1 and 2, we replicate earlier findings
with adults and children showing that listeners expect speakers to a)
produce informative utterances (Experiment 1) and b) communicate about
things that are relevant to common ground (Experiment 2). Based on these
results, we derive model predictions about how these two components
should be integrated. In Experiment 3, we test how listeners integrate
expectations and compare model predictions to empirical data. Experiment
4 replicates and extends Experiment 3 by gradually varying the strength
of common ground assumptions.

@@

Each experiment was implemented on a website (using JavaScript and HTML)
to which participants were directed. Figure \ref{fig:design} provides a
schematic overview of the setup and experimental procedures. The
instructions informed participants that they will see a number of animal
characters asking for novel toys. Participants' task was to identify the
toy a given animal is requesting. The basic layout involved two tables
with toys on them, located left and right of a little hill, on which the
animal was standing. For each animal, we recorded a set of utterances
(one speaker per animal) that were used throughout the experiments to
provide information and make requests. At test, toys were requested
using the following utterance: ``Oh cool, there is a {[}non-word{]} on
the table, how neat, can you give me the {[}non-word{]}?''. Participants
responded by clicking on one of the toys. Each experiment started with
two training trials in which animals requested familiar objects (car and
ball).

For all experiments, we pre-registered the sample size, experimental
design and the statistical analysis. For Experiment 3 and 4, we also
registered the model structure and predictions (see
\url{https://osf.io/u7kxe})

@@

\caption[Schematic experimental procedure]{Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by, or new to the speaker. Experiment 3 (right) combined manipulations. When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporally enlarged. Condition shown here: preference - same speaker - incongruent.}\label{fig:design}
\end{figure*}
\end{CodeChunk}

@@

All participants were recruited from Amazon Mechanical Turk and had US
IP addresses. We planned the sample size in each experiment to be 120
data points per cell. Experiment 1 had 40 participants.Informativeness
was manipulated by making one object less frequent in context. In the
test condition, one table contained one object of type A and the other
table contained one object of type A and one of type B (see Figure 1 in
red). On each trial, the animal introduced themselves, turned towards
the table with the two objects and made a request. If listeners expect
speakers to produce informative utterances, they should select object B.
This choice follows from the counterfactual inference that if the
(informative) speaker would have wanted to request A, they would have
turned to the table that only contained A. On the other hand, since B is
only located on the table together with A, there was no alternative way
to request B in a less ambiguous way. In the control condition, both
tables contained two objects, one of which was randomly determined as
the correct one. No inference was therefore licenced based on the
animals turning. Each participant received three trials in each
condition.

@@

Participants selected the less frequent object above chance in the test
condition (t = 5.51, \emph{p} \textless{} .001, see Figure
\ref{fig:plotexp12}) and did so more often compared to the control
condition (genreralized linear mixed model
(GLMM\footnote{All models included random effects for subject and speaker (animal).}):\emph{\(\beta\)}
= 1.49, se = 0.5 \emph{p} = .003). This result replicates earlier work
(Frank \& Goodman, 2014).

@@

\caption[Results from Experiment 1 and 2.For preference and novelty, control refers to a different speaker (see Fig]{Results from Experiment 1 and 2.For preference and novelty, control refers to a different speaker (see Fig. 1). Transparent dots show data from individual particpants, solid dots represent condition means, error bars are 95 \% CIs. Dashed line indicates performance expected by chance.}\label{fig:plotexp12}
\end{figure}
\end{CodeChunk}

@@

We manipulated common ground expectations based on two ways which have
succesfully been used in developmental studies (e.g. Akhtar et al.,
1996; Saylor et al., 2009). Speakers either expressed preference for one
object or one object was new to the speaker.

@@

There were 80 participants, 40 per manipulation. In the preference
manipulation, each table had a different object. In the beginning, the
animal appeared on the hill and introduced themselves. Next, they turned
to one of the tables and expressed either that they liked (``Oh wow, I
really like that one'') or disliked (``Oh bleh, I really don't like that
one'') the object before turning to the other side and expressing the
respective other attitude. Then, the animal disappeared. After a short
period of time, either the same or a different animal appeared and
requested an object while facing straight (see Figure 1 in blue). If
participants took into account the information they gained about the
speaker, they should select the previously preferred object if the
returning animal was the same. If a different animal returned, they
should choose randomly between objects.

In the novelty manipulation, one table was initially empty while there
was an object on the other one (see Figure \ref{fig:design}). The animal
turned to one of the sides and commented either on the presence (``Aha,
look at that'') or the absence of an object (``Hm, nothing there'').
Next, the animal quickly disappeared. The same animal re-appeared and
the sequence above was repeated. When the animal disappeared for the
second time, a second object appeared on the empty table while the
animal was away. Like in preference, we now manipulated if the same
animal or a different one would return. In case of the same animal
returning, listeners could infer the referent of the subsequent request
by considering that one object was new to the speaker and therefore more
likely to be of interest to them. No such inference was licensed when a
different animal returned because both objects were equally new to them.

@@

Participants selected the preferred object above chance when the same
animal returned (t = 29.14, \emph{p} \textless{} .001, see Figure
\ref{fig:plotexp12}) and did so more often compared to trials in which a
different animal returned (GLMM: \(\beta\) = 2.97, se = 0.7, \emph{p}
\textless{} .001). Thus, listeners inferred the referent of the
utterance by considering previous interactions with the speaker.
Interestingly, participants to some extend transferred preference from
one animal to the other and also selected the preferred object above
chance when a different animal returned (t = 2.7, \emph{p} = .01). In
sum, this study shows that adults make comparable infernces to children
(Saylor et al., 2009).

The novel object was selected above chance when the same animal returned
(t = 6.77, \emph{p} \textless{} .001) but not when a different one
appeared (t = 1.49, \emph{p} = .144, see Figure \ref{fig:plotexp12}).
Furthermore, the two conditions differed in the expected direction
(\(\beta\) = 1.7, se = 0.39, \emph{p} \textless{} .001). Thus, like
children (Akhtar et al., 1996), participants used their prior
information about the speaker to resolve ambiguity in their utterance.

@@

A total of 121 individuals participated in the experiment. The test
situation was the same as in the test condition in Experiment 1: One
table with object of type A and the other with an object of type A and
B. Again, the animal always turned to the table with two objects and
ambiguously requested an object. In Experiment 1, the listener had no
prior information about each object. In Experiment 3, however, we
manipulated common ground expectations in the same way as in Experiment
2. For example, the animal would turn to the table with one object and
express their dis-liking of object A, then they would turn to the other
table and express their liking of object B (not using labels). Next,
after quickly disappearing, the animal would reappear, turn to the table
with two objects and make a request.

For each version of Experiment 2, there were 4 conditions in Experiment
3. If the preferred/novel object was the less frequent one (object B),
the two expectations were congruent. If the preferred/novel object was
the more frequent one (object A), expectations were incongruent. For
each type of expectation alignment, we varied if the same or a different
animal returned. Participants either completed the preference or novelty
version and received two test trials in each of the four conditions.
Before discussing the empirical results, we briefly discuss the
cognitive model we used to predict expectation integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates
a pragmatic listener reasoning about a cooperative speaker who is trying
to refer to an object (Frank \& Goodman, 2012). The speaker chooses how
to refer to the object by reasoning about a naive listener who does not
know the labels for the object (Frank \& Goodman, 2014). Mathematically,
the probability that the listener chooses a referent given an utterance
is defined as follows:\\
\[P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)\] Here, \(P_S(u|r_s)\) is the
likelihood that the speaker will use an utterance to refer to a specific
referent. It is defined as the surprisal of an utterance given the
referent for a naive listener, who interprets utterances in a literal
way:

\[P_S(u|r_s)\propto exp(\alpha U_S(u;s))\]

The numerical strength of the expression above depends on \(\alpha\),
which can be interpreted as an indicator of how rational the speaker is
in choosing utterances.

@@

In our model, we used the results from Experiment 1 and 2 to specify
\(\alpha\) as well as \(P_S(r_s)\). The former was set to yield the
proportion measured in Experiment 1 in a model with uniform priors. The
prior probability for each object was the proportion with which this
object was chosen in Experiment
2\footnote{Proportions were measured when participants chose between two objects. However, in Experiment 3, three objects were involved. For each object we used the proportion measured in Experiment 2 as the prior probability.This approached spread out the absolute probability mass for each object but conserved the relative relation between objects.}.
Based on these parameter settings, we predicted the proportion with
which listeners will choose the less frequent object in each unique
combination of alignment and speaker identity separately for each prior
manipulation (eight conditions total, see Figure \ref{fig:plotexp3}). We
compared the fit of this pragmatic model to two alternative models using
Bayes Factors (BF). The first alternative model ignored the speaker
specific expectations (flat prior model) while the second ignored the
informativeness inference (prior only model). To capture the idea that
behavioral data is to some degree noisy, all models included a ``noise
parameter'', which reflects that, with a certain probability,
participants respond randomly instead of in line with the intended
manipulation on a given trial. Noise parameters are estimated based on
the data.

@@

@@

\caption[(A) Model predictions compared to data for Experiment 3 and (B) Experiment 4]{(A) Model predictions compared to data for Experiment 3 and (B) Experiment 4. (C) Data for strong prior manipualtion in Experiment 3 and 4. Error bars = 95 \%  HDIs.}\label{fig:plotmodelcomp}
\end{figure*}
\end{CodeChunk}

Results are discussed in the form of the proportion with which listeners
chose the less frequent object. For a comparison to chance within each
unique condition see Figure \ref{fig:plotexp3}. Both alignment and
speaker identity influenced participants' responses (GLMM model term:
\texttt{alignment*speaker}; \(\beta\) = -2.92, se = 0.57, \emph{p}
\textless{} .001). Figure \ref{fig:plotmodelcomp} shows the mean
response in each unique condition compared to the model. Model
predictions and data were highly correlated (r = 0.96, \emph{p}
\textless{} .001). Model fit was better in the model taking into account
both types of expectations compared to the flat prior (BF = 2e+79) or
prior only model (BF = 1.8e+34).

Interestingly, like in Experiment 2, there was a transfer of preference
in case of a speaker change. Participants were at chance in the
preference - different speaker - incongruent condition (see Figure
\ref{fig:plotexp3}). If preference would have been specific to a
particular individual, participants should have selected the less
frequent object above chance (as they did in the corresponding condition
with the novelty manipulation).

In Experiment 4, we replicated and extended Experiment 3 by manipulating
the strength of the common ground expectations. Our main goal was to
investigate the scope of our cognitive model and we therefore limit the
description of the methods and the discussion of the results to the
aspects relevant to the comparison between model predictions and data.

This experiment had 453 participants. The structure of the experiment
was the same as in Experiment 3. For each common ground expectation
(preference and novelty), we intended to have a strong, a medium and a
weak manipulation. The strength of each manipulation was determined by
the proportion with which participants chose the preferred / novel
object given the manipulation. We succeeded in this for novelty. For
preference we did not find a manipulation that yielded only a weak
preference.

The strong manipulations were identical to Experiment 3 and the results
are therefore a direct replication. For novelty, in the medium
manipulation, the animal turned to each table only once before the test.
In the weak manipulation, the animal only turned to the table with an
object before the test. In the medium manipulation for preference, the
animal only expressed liking and did so in a more subtle way (saying
only: ``Oh, wow''). Participants were assigned to one level of common
ground expectation and completed two test trials in each of the four
conditions.

Model predictions were obtained in the same way as in Experiment 3. The
likelihood term of the model was identical. For the each level of common
ground expectation, we measured the proportion with which participants
chose the preferred / novel object in a separate experiment. Due to
space limitations, we do not discuss them in any more detail here. Like
in Experiment 3, the results of these experiments were used to specify
the prior distribution in the model.

As noted above, the strong prior manipulation was a direct replication
of Experiment 3. Results from the two rounds of data collection were
highly correlated (r = 0.97, \emph{p} \textless{} .001, see Figure
\ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data
from Experiment 4 were highly correlated to the corresponding model
predictions (r = 0.91, \emph{p} \textless{} .001, see Figure
\ref{fig:plotmodelcomp}B). Again, the pragmatic model provided a better
fit compared to the flat prior (BF = 4.4e+74 ) or prior only model (BF =
1.8e+84).

@@

Language use and learning requires balancing different types of
expectations about one's interlocutor. Here we used a Bayesian cognitive
model to predict this integration process. Experiment 1 and 2 replicated
earlier studies showing that adult listeners expect speakers to produce
utterances in an informative way as well as in light of common ground.
Next, we combined the procedures from the first two experiments to study
how listeners would integrate them. We used the results from Experiment
1 and 2 to specify the model parameters that represented the two types
of expectations and make parameter-free predictions about new behavior.
Experiments 3 and 4 showed that both types of expectations influenced
listeners inferences. In general, listener behavior accurately described
by our model, suggesting that -- at least in population aggregate --
listeners trade off flexibly between speaker specific and general
pragmatic expectations.

>>>>

\begin{abstract}
Pragmatic inferences are an integral part of language learning and
comprehension. To recover the intended meaning of an utterance,
listeners need to balance and integrate different sources of contextual
information. In a series of experiments, we studied how listeners
integrate general expectations about speakers with expectations specific
to their interactional history with a particular speaker. We used a
Bayesian pragmatics model to formalize the integration process. In
Experiments 1 and 2, we replicated previous findings showing that
listeners make inferences based on speaker-general and speaker-specific
expectations. We then used the empirical measurements from these
experiments to generate model predictions about how the two kinds of
expectations should be integrated, which we tested in Experiment 3.
Experiment 4 replicated and extended Experiment 3 to a broader set of
conditions. In both experiments, listeners based their inferences on
both types of expectations. We found that model performance was also
consistent with this finding; with better fit for a model which
incorporated both general- and specific-information compared to
baselines incorporating only one type. Listeners flexibly integrate
different forms of social expectations across a range of contexts, a
process which can be described using Bayesian models of pragmatic
reasoning. Our approach highlights how experiments and computational
models can be used in conjunction to inform theories about language use
and comprehension.

@@

One of the most astonishing features of natural language is that it
allows us to communicate precise meanings despite the fact that most
utterances are inherently ambiguous. While the conventional mapping
between sounds (words) and objects constrain what a speaker may mean by
an utterance, the intended meaning of the utterance is not reducible to
the words that are contained in it. It takes additional pragmatic
inference to recover the intended meaning (Levinson, 2000).

Pragmatic inferences rest on a set of expectations that interlocutors
bring to the table when entering a communicative interaction. On the one
hand, speakers and listeners have the general expectation that their
partner communicates in an informative and relevant way (Sperber \&
Wilson, 2001). Grice (1991) summarised this expectation via the
``Cooperative Principle'': ``Make your contribution such as is required,
at the stage at which it occurs, by the accepted purpose or direction of
the talk exchange in which you are engaged.'' Importantly, the second
half of the Cooperative Principle describes a second type of
expectation: interlocutors expect each other to produce and interpret
utterances \emph{in light of the shared common ground between them} (H.
H. Clark, 1996). Common ground refers to bits of information that are
assumed to be shared, either because they were mentioned over the course
of the conversation or grounded through some form of joint experience
(Bohn \& Koymen, 2018). Note that by its very nature, common ground may
vary with the individuals involved in a particular conversation.

These same general and specific expectations can support children's word
learning (E. V. Clark, 2009; Tomasello, 2009). On the one hand, children
have been found to learn novel words by assuming speakers are generally
informative (Frank \& Goodman, 2014). That is, in the absence of any
prior interaction with the speaker, children interpreted a novel word as
referring to the most informative referent. On the other hand, children
use conversation-specific common ground expectations to decide which
object a specific speaker is referring to when they use a novel word
(Akhtar, Carpenter, \& Tomasello, 1996). For example, when a speaker
expressed preference for a particular object, children expect a novel
word from the same speaker to refer to the previously preferred object
(Saylor, Sabbagh, Fortuna, \& Troseth, 2009).

But how do listeners integrate general and common ground-related
expectations during word learning? Are pragmatic inferences strengthened
additively when both support a particular interpretation? How are they
weighed when they are in conflict? The Rational Speech Act (RSA)
framework (Frank \& Goodman, 2012; Goodman \& Frank, 2016) offers a
formal framework for addressing this information integration problem.
RSA models are characterized by a recursive structure in which a
pragmatic listener tries to uncover a speaker's intended meaning by
assuming the speaker chose their utterance in order to get a naive
listener to recover their intended meaning. RSA models have made
accurate quantitative predictions about various forms of pragmatic
language use and word learning (Goodman \& Frank, 2016). However, a
comprehensive treatment of how general and common ground expectations
are integrated is still missing.

Within RSA models, each agent in this recursion is modeled as a Bayesian
reasoner; thus, information integration is treated as a process of
probabilistic inference. The speaker-general informativeness expectation
is already encoded in the structure of the model: Speakers produce
utterances to aid the listener in disambiguating referents. We
operationalize speaker-specific, common ground information as the shared
prior probability of referents in the context of the utterance. Thus, a
natural locus for information integration within these models is the
tradeoff between the prior probability of a particular referent and the
likelihood of that referent given the current utterance.

Here we evaluate this rational, pragmatic account of information
integration. We isolate speaker-specific and common-ground information
experimentally, then test how adult listeners combine them in a word
learning setting. In Experiments 1 and 2, we replicate findings showing
that listeners expect speakers to a) produce informative utterances
(Experiment 1) and b) communicate about things that are relevant to
common ground (Experiment 2). Based on these results, we generate model
predictions using the RSA framework about how these two components
should be integrated. In Experiment 3, we test how listeners integrate
their expectations and compare model predictions to empirical data.
Experiment 4 replicates and extends Experiment 3 by varying the strength
of common ground assumptions.

@@

Experiments were conducted online using Amazon's Mechanical Turk. Figure
\ref{fig:design} provides a schematic overview of the setup and
experimental procedures. The instructions informed participants that
they would see a number of animal characters asking for novel toys.
Participants were asked to identify the toy being requested by a
particular animal. The basic layout involved two tables with toys on
them, located left and right of a little hill, on which the animal was
standing. For each animal, we recorded a set of utterances (one speaker
per animal) that were used throughout the experiments to provide
information and make requests. At test, toys were requested using the
following utterance: ``Oh cool, there is a {[}non-word{]} on the table,
how neat, can you give me the {[}non-word{]}?''. Participants responded
by clicking on one of the toys. Each experiment started with two
training trials in which animals requested familiar objects (car and
ball).

For all experiments, we pre-registered the sample size, experimental
design and the statistical analysis. For Experiment 3 and 4, we also
registered the model structure and predictions (see {[}masked for peer
review{]})

@@

\caption[Schematic experimental procedure]{Schematic experimental procedure. In all conditions, at test (bottom), the speaker ambiguously requested an object using a non-word (e.g. “dax”). Participants clicked on the object they thought the speaker referred to. Informativeness (Experiment 1, left) translated to making one object less frequent in context. Common ground (Experiment 2, middle) was manipulated by making one object prefered by or new to the speaker. Experiment 3 (right) combined manipulations. When expressing e.g. preference for an object on a table with two objects (panel 3 from top), the respective object was temporarily enlarged. Condition shown here: preference - same speaker - incongruent. Green plus signs represent an expression of preference and red minus of dispreference (see main text for details).}\label{fig:design}
\end{figure*}
\end{CodeChunk}

@@

All participants were recruited from Amazon Mechanical Turk and had US
IP addresses. Sample size in each experiment was planned to be 120 data
points per cell. Experiment 1 had 40 participants. In the test
condition, one table contained one object of type A and the other table
contained one object of type A and one of type B (see Figure
\ref{fig:design}, left). On each trial, the animal introduced themselves
(e.g. ``Hi, I'm Dog''), turned towards the table with the two objects
and made a request. If listeners expect speakers to produce informative
utterances, they should select object B. This choice follows from the
counterfactual inference that if the (informative) speaker would have
wanted to request A, they would have turned to the table that only
contained A. On the other hand, since B is only located on the table
together with A, there was no alternative way to request B in a less
ambiguous way. In the control condition, both tables contained two
objects, one of which was randomly determined as the correct one. No
inference was therefore licenced. Each participant received three trials
in each condition for a total of six trials, presented in a randomized
order.

@@

Participants selected the less frequent object above chance in the test
condition (t(39) = 5.51, \emph{p} \textless{} .001, see Figure
\ref{fig:plotexp12}) and did so more often compared to the control
condition (genreralized linear mixed model
(GLMM\footnote{All models included random effects for subject and speaker (animal).}):\emph{\(\beta\)}
= 1.49, se = 0.5 \emph{p} = .003). This result replicates earlier work
(Frank \& Goodman, 2014) and is consistent with the hypothesis that
listeners expect speakers to communicate in an informative way.

@@

\caption[Results from Experiment 1 and 2]{Results from Experiment 1 and 2. For preference and novelty, control refers to a different speaker (see Fig. 1). Transparent dots show data from individual participants, solid dots represent condition means, error bars are 95 \% CIs. Dashed line indicates performance expected by chance.}\label{fig:plotexp12}
\end{figure}
\end{CodeChunk}

@@

We manipulated common ground expectations based on procedures that have
successfully been used in developmental studies (e.g. Akhtar et al.,
1996; Saylor et al., 2009). Speakers either expressed preference for one
object or one object was new to the speaker.

@@

We collected data from 80 participants, with 40 in each condition. In
the preference condition, each table had a different object. In the
beginning, the animal appeared on the hill and introduced themselves.
Next, they turned to one of the tables and expressed either that they
liked (``Oh wow, I really like that one'') or disliked (``Oh bleh, I
really don't like that one'') the object before turning to the other
side and expressing the respective other attitude. Then, the animal
disappeared. After a short period of time, either the same or a
different animal appeared and requested an object while facing straight
ahead (see Figure \ref{fig:design}, middle). If participants took into
account the information they gained about the speaker, they should
select the previously preferred object if the returning animal was the
same. If a different animal returned, they should choose randomly
between objects.

In the novelty condition, one table was initially empty while there was
an object on the other table (see Figure \ref{fig:design}). The animal
turned to one of the sides and commented either on the presence (``Aha,
look at that'') or the absence of an object (``Hm, nothing there'').
Next, the animal disappeared. The same animal re-appeared and the
sequence above was repeated. When the animal disappeared for the second
time, a second object appeared on the empty table while the animal was
away. Like in the preference condition, we now manipulated if the same
animal or a different one returned. In case of the same animal
returning, listeners could infer the referent of the subsequent request
by considering that one object was new to the speaker and therefore more
likely to be of interest to them. However, no such inference was
licensed when a different animal returned because both objects were
novel.

@@

Participants selected the preferred object above chance when the same
animal returned (t(39) = 29.14, \emph{p} \textless{} .001, see Figure
\ref{fig:plotexp12}) and did so more often compared to trials in which a
different animal returned (GLMM: \(\beta\) = 2.97, se = 0.7, \emph{p}
\textless{} .001). Thus, listeners inferred the referent of the
utterance by considering previous interactions with the speaker.
Interestingly, participants transferred preference to some extent from
one animal to the other and selected the preferred object above chance
when a different animal returned (t(39) = 2.7, \emph{p} = .01). In sum,
this study shows that adults make comparable inferences to children
(Saylor et al., 2009).

The novel object was selected above chance when the same animal returned
(t(39) = 6.77, \emph{p} \textless{} .001) but not when a different one
appeared (t(39) = 1.49, \emph{p} = .144, see Figure
\ref{fig:plotexp12}). Furthermore, the two conditions differed in the
expected direction (\(\beta\) = 1.7, se = 0.39, \emph{p} \textless{}
.001). Thus, like children (Akhtar et al., 1996), adults used their
prior information about the speaker to resolve ambiguity in the
utterance.

@@

A total of 121 individuals participated in the experiment. The test
situation was the same as in the test condition in Experiment 1: One
table with object of type A and the other with an object of type A and
B. Again, the animal always turned to the table with two objects and
ambiguously requested an object. In Experiment 1, the listener had no
prior information about each object. In Experiment 3, however, we
manipulated common ground expectations in the same way as in Experiment
2. For example, the animal would turn to the table with one object and
express that they don't like object A, then they would turn to the other
table and express that they like object B. Next, after quickly
disappearing, the animal would reappear, turn to the table with two
objects and make a request (see Figure \ref{fig:design}, right).

For each common ground condition, there were 4 conditions in Experiment
3 resulting from the cross of congruent/incongruent informativeness with
same/different speaker. If the preferred/novel object was the less
frequent one (object B), the two expectations were congruent. If the
preferred/novel object was the more frequent one (object A),
expectations were incongruent. For each type of expectation alignment,
we varied if the same or a different animal returned. Participants
either completed the preference or novelty version with two test trials
in each of the four conditions. Before discussing the empirical results,
we briefly discuss the cognitive model we used to predict expectation
integration.

@@

To derive predictions, we used a probabilistic RSA model that simulates
a pragmatic listener reasoning about a cooperative speaker who is trying
to refer to an object (Frank \& Goodman, 2012). The speaker chooses how
to refer to the object by reasoning about a naive listener who does not
know the labels for the object (Frank \& Goodman, 2014). The conditional
probability that the listener chooses a referent given an utterance is
defined as follows: \[P_L(r_s|u)\propto P_S(u|r_s)P_S(r_s)\] Here,
\(P_S(u|r_s)\) is the likelihood that the speaker will use an utterance
u to refer to a specific referent r. It is defined in terms of a utility
function \(U_S(u;s)\) consisting of the surprisal of \(u\) for a naive
listener \(L_0\), who interprets \(u\) according its literal semantics:
\[P_S(u|r_s)\propto exp(\alpha U_S(u;s))\] The numerical strength of the
expression above depends on \(\alpha\) scalar value, α, which can be
interpreted as an indicator of how rational the speaker is in choosing
utterances (i.e.~as alpha increases, the speaker is more likely to
choose the most informative utterance).

@@

We used the results from Experiment 1 and 2 to specify \(\alpha\) as
well as \(P_S(r_s)\) in our model. We set \(\alpha\) so that a model
with uniform priors would predict the average proportion measured in
Experiment 1. The prior probability for each object was set to be the
proportion with which this object was chosen in Experiment
2\footnote{Proportions were measured when participants chose between two objects. However, in Experiment 3, three objects were involved. For each object we used the proportion measured in Experiment 2 as the prior probability.This approached spread out the absolute probability mass for each object but conserved the relative relation between objects.}.
Based on these parameter settings, we predicted the proportion with
which listeners will choose the more informative object in each of eight
unique conditions mentioned above (see also Figure \ref{fig:plotexp3}).
We compared the fit of this pragmatic model to two alternative models
using Bayes Factors (BF). The first alternative model ignored the
speaker specific expectations (uniform prior model) while the second
ignored the informativeness inference (prior only model).

@@

Results are discussed in the form of the proportion with which listeners
chose the more informative object (i.e., the object that would be the
more informative referent when only considering speaker general
information). For a comparison to chance within each unique condition
see \ref{fig:plotexp3}. Combinations of alignment and speaker identity
differed in how they influenced participants' responses (GLMM model
term: \texttt{alignment*speaker}; \(\beta\) = -2.92, se = 0.57, \emph{p}
\textless{} .001). Figure \ref{fig:plotmodelcomp}A and B shows the mean
response in each unique condition compared to the pragmatics model.
Model predictions and data were highly correlated (r = 0.96, \emph{p}
\textless{} .001).Model fit was much better in the model taking into
account both types of expectations compared to the uniform prior (BF =
2e+79) or prior only model (BF = 1.8e+34).

Interestingly, as in Experiment 2, there was a transfer of preference in
the case of speaker change. Participants were at chance in the
preference - different speaker - incongruent condition (see Figure
\ref{fig:plotexp3}). If preference would have been specific to a
particular individual, participants should have selected the less
frequent object above chance (as they did in the corresponding condition
with the novelty manipulation). Because it takes into account the
measurement from the earlier experiment, our model predicts these
results; future work might explicitly model generalization across
speakers.

In Experiment 4, we replicated and extended Experiment 3 by manipulating
the strength of the common ground expectations. Our main goal was to
investigate the quantitative accuracy of our cognitive model and we
therefore limit the description of the methods and the discussion of the
results to the aspects relevant to the comparison between model
predictions and data.

This experiment had 453 participants. The structure of the experiment
was the same as in Experiment 3. For each common ground expectation
(preference and novelty), we intended to have a strong, a medium and a
weak condition. The strength of each condition was determined by the
proportion with which participants chose the preferred/novel object
given the manipulation. We succeeded in generating quantitative
variability for novelty. For preference we piloted a number of
additional manipulations but did no find one that yielded a weaker
preference compared to a medium condition.

The strong manipulations were identical to Experiment 3 and the results
are therefore a direct replication (see Figure
\ref{fig:plotmodelcomp}C). For novelty, in the medium condition, the
animal turned to each table only once before the test. In the weak
condition, the animal only turned to the table with an object before the
test (instead of turning to and commenting on both). In the medium
condition for preference, the animal only expressed liking and did so in
a more subtle way (saying only: ``Oh, wow'' while pointing to the
object). Participants were assigned to one level of common ground
expectation and completed two test trials in each of the four conditions
(alignment x speaker change).

Model predictions were obtained in the same way as in Experiment 3; with
\(\alpha\) inferred from the data and \(P_S(r_s)\) measured empirically
(in a set of corresponding experiments parallel to Experiment 2).

@@

\caption[(A) Model predictions compared to data for Experiment 3 and (B) Experiment 4]{(A) Model predictions compared to data for Experiment 3 and (B) Experiment 4. (C) Data for strong prior manipulation in Experiment 3 and 4, providing a noise ceiling for the reliability of measurements. Error bars = 95 \%  HDIs.}\label{fig:plotmodelcomp}
\end{figure*}
\end{CodeChunk}

As noted above, the strong prior condition was a direct replication of
Experiment 3. Results from the two rounds of data collection were highly
correlated (r = 0.97, \emph{p} \textless{} .001, see Figure
\ref{fig:plotmodelcomp}C). Across levels of prior manipulation, the data
from Experiment 4 were highly correlated to the corresponding model
predictions (r = 0.91, \emph{p} \textless{} .001, see Figure
\ref{fig:plotmodelcomp}B). Again, the pragmatics model provided a much
better fit compared to the flat prior (BF = 4.4e+74 ) or prior only
model (BF = 1.8e+84).

@@

Language use and learning requires balancing different types of
expectations about one's interlocutor - both expectations about how
speakers behave in general and specific expectations about how a
particular speaker might behave in a particular context. Here we used a
computational pragmatics model to predict this integration process.
Experiment 1 and 2 replicated previous studies showing that adult
listeners expect speakers to produce utterances informatively and also
with respect to common ground. We then combined the procedures from the
first two experiments to study how listeners would integrate these two
types of information. We used the results from Experiment 1 and 2 to
specify model parameters that represented the two types of expectations,
generating predictions about new behavior. Experiments 3 and 4 showed
that both types of expectations influenced listeners inferences. In
general, listener behavior was accurately described by our model,
suggesting that listeners trade-off flexibly between speaker specific
and general pragmatic expectations.

Notably, Experiment 3 also included situations in which the two
expectations were in conflict. For example, in some trials the speaker
expressed preference for object A, which was also the more frequent
(less informative) object. In these situations, a majority of
participants chose the preferred object as the referent (see also Figure
\ref{fig:plotexp3} preference--same speaker--incongruent). A simple
explanation for this pattern might be that common ground manipulations
were simply ``stronger'', corroborated by the fact they produced higher
rates of expected choice than the informativeness expectation when the
two were presented in isolation (see Figure \ref{fig:plotexp12}). In
Experiment 4, however, the medium manipulation for novelty yielded
numerically weaker results compared to the informativeness expectation
in Experiment 1, and yet participants still selected the novel object
above chance when the expectations were in conflict. Why is this?
